\pkg{cubble}: An R Package for Structuring Spatio-temporal Data

### 2022-03-31 2-0-cube.Rmd
draft on conceptual cube and new challenges

# Conceptual framework

Multivariate spatio-temporal data can be conceptualised using a cubical data model with reference to variables, time and space. This naturally motivates the usage of array to represent and store spatio-temporal data, as evident by satellite imageries, large climate models, among others. The work by @lu_multidimensional_2018 has discussed the suitability of this representation through array algebra and practical spatio-temporal tasks and software implementation has been seen in many R packages: \pkg{raster} [@raster], \pkg{terra} [@terra], and \pkg{stars} [@stars].

While it is efficient to represent spatio-temporal data in an array, there are situation where array [is not the best solution]. mutate in the array is not clearly define (how to add a column vector)

Also, array is not the best solution for a large subset of data, specifically, the vector spatio-temporal data. With vector data, spatial and temporal variables are commonly collected in separate 2D tables and there is usually a location ID to link the spatial and temporal information. Sometimes, the geometrical shape of the location may not present and an additional linking is needed to incorporate the map data. With the multiple pieces in 2D data frame format, there should be a solution to organise all these data from multiple sources in a 2D structure, without needing to convert it into higher dimensional representation.

Reliable software and data structure has been proposed in the spatial and temporal domain, while a direct extension to spatio-temporal data is not ideal. The main issue here lays in the format that spatial and temporal data is stored. For spatial data, \pkg{sf} [@sf] covers a wide range of geometrical operations and is the format most map data stored. When incorporating the time dimension, a time wide format, introduced in @spacetime, is usually adopted. This treats time as variable name rather than value and makes it inconvenient to wrangle time, as per the tidy data principle [@tidydata]. In the temporal domain, \pkg{tsibble} [@tsibble] follows the tidy data principle to arrange the time in the long format, while when adding spatial variables, a long format implies repeating variables at each time and this can make a huge difference when the repetition is on large geometrical shapes with long time period.

>>>>

# Conceptual framework: spatio-temporal cube

Multivariate spatio-temporal data can be conceptualised using a cubical data model with reference to variables, time and space. This naturally motivates the usage of array to represent and store spatio-temporal data, as evident by satellite imageries, large climate models, among others. The work by @lu_multidimensional_2018 has discussed the suitability of this representation through array algebra and practical spatio-temporal tasks.

The cubical conceptual framework provides a generalisation to the operation can be done on the data for visualisation. A paper by @bach_review_2014 has reviewed the temporal data visualisation based on space-time cube operations. Notice that the term space-time cube in their article "does not need to involve spatial data", but referring to "an abstract 2D substrate that is used to visualize data at a specific time". Despite its main focus is on temporal data, the mindset of abstracting out data representation to construct visualisation is applicable to our spatio-temporal data manipulation and visualisation.

<!-- and software implementation has been seen in many R packages: \pkg{raster} [@raster], \pkg{terra} [@terra], and \pkg{stars} [@stars]. -->

# Existing work and new challenges

**what has been done and the new challenges**

The \pkg{spacetime} paper [@spacetime] has set the foundation for wrangling spatio-temporal data in \proglang{R}. It proposes four layouts (STF: full grid layout, STS: sparse grid layout, STI: irregular layout, and STT: trajectory) to represent spatio-temporal data. The implementation uses class \pkg{sp} to represent spatial data and class \pkg{xts} for temporal data. In recent years, new data structures have appeared to replace and improve on these two classes to represent spatial and temporal data. The package \pkg{sf} has adopted the simple feature access [@jr_herring_opengis_2011] to represent spatial information using POINT, MULTIPOINT, LINESTRING, POLYGON, MULTILINESTRING, MULTIPOLYGON, and GEOMETRYCOLLECTION. For temporal data, the package \pkg{tsibble} adopts the long form to explicitly record the time stamp as a column rather than an attribute of the data. These changes in the spatial and temporal community have yet been reflected in the spatio-temporal data analysis. The @spacetime paper also summarises three formats spatio-temporal data are usually recorded: time-wide, space-wide, and long format. With the concept of tidy data [@tidydata], the long form has been increasingly considered as the tidy format. The main advantages of it over the wide format is that the latter treat either time or location value as variable names (Section 3.1 in @tidydata), making them difficult to be wrangled. While temporal data can be benefited from a long format, it brings repetition of spatial variables in the combined spatio-temporal data and this can be a huge inefficiency for large MULTIPOLOYGON over long time period. This suggests the need to look for other format for spatio-temporal data.

<!-- For spatial data, \pkg{sf} [@sf] covers a wide range of geometrical operations and is the format most map data stored. When incorporating the time dimension, a time wide format, introduced in @spacetime, is usually adopted. This treats time as variable name rather than value and makes it inconvenient to wrangle time, as per the tidy data principle [@tidydata]. In the temporal domain, \pkg{tsibble} [@tsibble] follows the tidy data principle to arrange the time in the long format, while when adding spatial variables, a long format implies repeating variables at each time and this can make a huge difference when the repetition is on large geometrical shapes with long time period.  -->

**two challenages: multiple data sources + large spatio-temporal object**
In additional to this, there are more challenges in manipulation spatio-temporal data. Different variables are recorded and collected in multiple tables, for example, an areal data would map data for a collection of areas, geo-scientific data would record locations and other meta data, and the temporal variablem indexed by both location and time are stored in the long form table. To work with these data, analysts need to link the geo-scientific data with the temporal variables and work out how they are contained in the area. This would require merging all three sources together and can often go wrong because of the slightest unmatched of linking variable in different sources. Another challenge analysts face when wrangling spatio-temporal data is with the simple feature access representation of spatial data. While they have become the standard for recording large and complex spatial geometry and precise operation on the spatial coordinates, with data analysis, the detail is more than enough. A data analysts don't necessarily need to have the every nitty gritty detail of the location and a map with sufficient information to outline the shape of the location would be enough for the propose of visualisation. With a large and detailed geometrical object, the map would take longer to render and analysts always need to first simplify it into a more manageable size. Also, while the simple feature access structure the coordinates into a list column, analysts would sometimes like to unpack it into two columns of longitude and latitude to work with. They should be provided with the option to do so.

**A newer implementation**
The \pkg{stars} has been an implementation to represent spatio-temporal data with array. While it is efficient to represent spatio-temporal data in an array, there are situation where array [is not the best solution]. mutate in the array is not clearly define (how to add a column vector). Also, it doesn't solve the problem with multiple data sources as it may take a large amount of effort to shape multivariate spatio-temporal data into high-dimensional cube before the analysis starts and they should be provided with option to work with 2D tables.

<!-- Also, array is not the best solution for a large subset of data, specifically, the vector spatio-temporal data. With vector data, spatial and temporal variables are commonly collected in separate 2D tables and there is usually a location ID to link the spatial and temporal information. Sometimes, the geometrical shape of the location may not present and an additional linking is needed to incorporate the map data. With the multiple pieces in 2D data frame format, there should be a solution to organise all these data from multiple sources in a 2D structure, without needing to convert it into higher dimensional representation. -->

### 2022-03-31 main.tex
draft on conceptual cube and new challenges

\hypertarget{conceptual-framework}{%
\section{Conceptual framework}\label{conceptual-framework}}

Multivariate spatio-temporal data can be conceptualised using a cubical
data model with reference to variables, time and space. This naturally
motivates the usage of array to represent and store spatio-temporal
data, as evident by satellite imageries, large climate models, among
others. The work by \citet{lu_multidimensional_2018} has discussed the
suitability of this representation through array algebra and practical
spatio-temporal tasks and software implementation has been seen in many
R packages: \pkg{raster} \citep{raster}, \pkg{terra} \citep{terra}, and
\pkg{stars} \citep{stars}.

While it is efficient to represent spatio-temporal data in an array,
there are situation where array {[}is not the best solution{]}.
Sometimes, one may want to create a new

Also, array is not the best solution for a large subset of data,
specifically, the vector spatio-temporal data. With vector data, spatial
and temporal variables are commonly collected in separate 2D tables and
there is usually a location ID to link the spatial and temporal
information. Sometimes, the geometrical shape of the location may not
present and an additional linking is needed to incorporate the map data.
With the multiple pieces in 2D data frame format, there should be a
solution to organise all these data from multiple sources in a 2D
structure, without needing to convert it into higher dimensional
representation.

Reliable software and data structure has been proposed in the spatial
and temporal domain, while a direct extension to spatio-temporal data is
not ideal. The main issue here lays in the format that spatial and
temporal data is stored. For spatial data, \pkg{sf} \citep{sf} covers a
wide range of geometrical operations and is the format most map data
stored. When incorporating the time dimension, a time wide format,
introduced in \citet{spacetime}, is usually adopted. This treats time as
variable name rather than value and makes it inconvenient to wrangle
time, as per the tidy data principle \citep{tidydata}. In the temporal
domain, \pkg{tsibble} \citep{tsibble} follows the tidy data principle to
arrange the time in the long format, while when adding spatial
variables, a long format implies repeating variables at each time and
this can make a huge difference when the repetition is on large
geometrical shapes with long time period.

However, with all the benefit of conceptualising the data into a cube,
the cubical structure is not particular convenient for, using the
tidyverse word, mutating new variables in the data object. This is
because an array is a higher dimensional object than the 2D data frame
and adding a new column doesn't clearly define how a vector should be
added to an (multidimensional) array.

The cubble package is a collection of tools to work with vector
spatio-temporal data from multiple sources.

>>>>

\hypertarget{conceptual-framework-spatio-temporal-cube}{%
\section{Conceptual framework: spatio-temporal
cube}\label{conceptual-framework-spatio-temporal-cube}}

Multivariate spatio-temporal data can be conceptualised using a cubical
data model with reference to variables, time and space. This naturally
motivates the usage of array to represent and store spatio-temporal
data, as evident by satellite imageries, large climate models, among
others. The work by \citet{lu_multidimensional_2018} has discussed the
suitability of this representation through array algebra and practical
spatio-temporal tasks.

The cubical conceptual framework provides a generalisation to the
operation can be done on the data for visualisation. A paper by
\citet{bach_review_2014} has reviewed the temporal data visualisation
based on space-time cube operations. Notice that the term space-time
cube in their article ``does not need to involve spatial data'', but
referring to ``an abstract 2D substrate that is used to visualize data
at a specific time''. Despite its main focus is on temporal data, the
mindset of abstracting out data representation to construct
visualisation is applicable to our spatio-temporal data manipulation and
visualisation.

\hypertarget{existing-work-and-new-challenges}{%
\section{Existing work and new
challenges}\label{existing-work-and-new-challenges}}

\textbf{what has been done and the new challenges}

The \pkg{spacetime} paper \citep{spacetime} has set the foundation for
wrangling spatio-temporal data in \proglang{R}. It proposes four layouts
(STF: full grid layout, STS: sparse grid layout, STI: irregular layout,
and STT: trajectory) to represent spatio-temporal data. The
implementation uses class \pkg{sp} to represent spatial data and class
\pkg{xts} for temporal data. In recent years, new data structures have
appeared to replace and improve on these two classes to represent
spatial and temporal data. The package \pkg{sf} has adopted the simple
feature access \citep{jr_herring_opengis_2011} to represent spatial
information using POINT, MULTIPOINT, LINESTRING, POLYGON,
MULTILINESTRING, MULTIPOLYGON, and GEOMETRYCOLLECTION. For temporal
data, the package \pkg{tsibble} adopts the long form to explicitly
record the time stamp as a column rather than an attribute of the data.
These changes in the spatial and temporal community have yet been
reflected in the spatio-temporal data analysis. The \citet{spacetime}
paper also summarises three formats spatio-temporal data are usually
recorded: time-wide, space-wide, and long format. With the concept of
tidy data \citep{tidydata}, the long form has been increasingly
considered as the tidy format. The main advantages of it over the wide
format is that the latter treat either time or location value as
variable names (Section 3.1 in \citet{tidydata}), making them difficult
to be wrangled. While temporal data can be benefited from a long format,
it brings repetition of spatial variables in the combined
spatio-temporal data and this can be a huge inefficiency for large
MULTIPOLOYGON over long time period. This suggests the need to look for
other format for spatio-temporal data.

\textbf{two challenages: multiple data sources + large spatio-temporal
object} In additional to this, there are more challenges in manipulation
spatio-temporal data. Different variables are recorded and collected in
multiple tables, for example, an areal data would map data for a
collection of areas, geo-scientific data would record locations and
other meta data, and the temporal variablem indexed by both location and
time are stored in the long form table. To work with these data,
analysts need to link the geo-scientific data with the temporal
variables and work out how they are contained in the area. This would
require merging all three sources together and can often go wrong
because of the slightest unmatched of linking variable in different
sources. Another challenge analysts face when wrangling spatio-temporal
data is with the simple feature access representation of spatial data.
While they have become the standard for recording large and complex
spatial geometry and precise operation on the spatial coordinates, with
data analysis, the detail is more than enough. A data analysts don't
necessarily need to have the every nitty gritty detail of the location
and a map with sufficient information to outline the shape of the
location would be enough for the propose of visualisation. With a large
and detailed geometrical object, the map would take longer to render and
analysts always need to first simplify it into a more manageable size.
Also, while the simple feature access structure the coordinates into a
list column, analysts would sometimes like to unpack it into two columns
of longitude and latitude to work with. They should be provided with the
option to do so.

\textbf{A newer implementation} The \pkg{stars} has been an
implementation to represent spatio-temporal data with array. While it is
efficient to represent spatio-temporal data in an array, there are
situation where array {[}is not the best solution{]}. mutate in the
array is not clearly define (how to add a column vector). Also, it
doesn't solve the problem with multiple data sources as it may take a
large amount of effort to shape multivariate spatio-temporal data into
high-dimensional cube before the analysis starts and they should be
provided with option to work with 2D tables.
